---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[`qdatatable`](https://github.com/WinVector/qdatatable) is an implementation of
the [`rquery`](https://github.com/WinVector/rquery) piped relational algebra 
hosted on [`data.tabl`](http://r-datatable.com).  `qdatatabl` allow the expression
of complex transformations as a series of relational operators.  For example 
scoring a logistic regression model (which requires grouping, ordering, and ranking)
is organized as follows.  For more on this example please see 
["Let’s Have Some Sympathy For The Part-time R User"](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).

```{r}
library("qdatatable")
library("data.table")
```


```{r}
# data example
dL <- wrapr::build_frame(
   "subjectID", "surveyCategory"     , "assessmentTotal" |
   1          , "withdrawal behavior", 5                 |
   1          , "positive re-framing", 2                 |
   2          , "withdrawal behavior", 3                 |
   2          , "positive re-framing", 4                 )
```


```{r}
scale <- 0.237

# example rquery pipeline
rquery_pipeline <- local_td(dL) %.>%
  extend_nse(.,
             one := 1) %.>%
  extend_nse(.,
             probability :=
               exp(assessmentTotal * scale)/
               sum(exp(assessmentTotal * scale)),
             count := sum(one),
             rank := rank(probability, surveyCategory),
             partitionby = 'subjectID') %.>%
  extend_nse(.,
             isdiagnosis := rank == count,
             diagnosis := surveyCategory) %.>%
  select_rows_nse(., 
                  isdiagnosis == TRUE) %.>%
  select_columns(., 
                 c('subjectID', 'diagnosis', 'probability')) %.>%
  orderby(., 'subjectID')
```

Show expanded form of query tree.

```{r, comment=""}
cat(format(rquery_pipeline))
```


```{r}
ex_data_table(rquery_pipeline)[]
```

Initial benchmarking shows `qdataframe` to be 4 times faster than `dplyr` (notes [here](https://github.com/WinVector/rquery/blob/master/extras/data_table.md)).

To install `qdatatable` please use `devtools` as follows.

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("WinVector/qdataframe")
```

