---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[`rqdatatable`](https://github.com/WinVector/rqdatatable) is an implementation of
the [`rquery`](https://github.com/WinVector/rquery) piped relational algebra 
hosted on [`data.table`](http://r-datatable.com).  `rquery` allow the expression
of complex transformations as a series of relational operators and
`rqdatatable` implements the operators using `data.table`.

For example 
scoring a logistic regression model (which requires grouping, ordering, and ranking)
is organized as follows.  For more on this example please see 
["Let’s Have Some Sympathy For The Part-time R User"](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).

```{r}
library("rqdatatable")
```


```{r}
# data example
dL <- build_frame(
   "subjectID", "surveyCategory"     , "assessmentTotal" |
   1          , "withdrawal behavior", 5                 |
   1          , "positive re-framing", 2                 |
   2          , "withdrawal behavior", 3                 |
   2          , "positive re-framing", 4                 )
```


```{r}
scale <- 0.237

# example rquery pipeline
rquery_pipeline <- local_td(dL) %.>%
  extend_nse(.,
             probability :=
               exp(assessmentTotal * scale))  %.>% 
  normalize_cols(.,
                 "probability",
                 partitionby = 'subjectID') %.>%
  pick_top_k(.,
             k = 1,
             partitionby = 'subjectID',
             orderby = c('probability', 'surveyCategory'),
             reverse = c('probability', 'surveyCategory')) %.>% 
  rename_columns(., c('diagnosis' = 'surveyCategory')) %.>%
  select_columns(., c('subjectID', 
                      'diagnosis', 
                      'probability')) %.>%
  orderby(., cols = 'subjectID')
```

We can show the expanded form of query tree.

```{r, comment=""}
cat(format(rquery_pipeline))
```

And execute it using `data.table`.

```{r}
ex_data_table(rquery_pipeline)[]
```

Can also execute on new tables.

```{r}
rquery_executor <- list(f = rqdatatable::ex_data_table)

build_frame(
   "subjectID", "surveyCategory"     , "assessmentTotal" |
   7          , "withdrawal behavior", 5                 |
   7          , "positive re-framing", 20                ) %.>%
  rquery_pipeline
```


Initial bench-marking of `rqdatatable` is very favorable (notes [here](http://www.win-vector.com/blog/2018/06/rqdatatable-rquery-powered-by-data-table/)).

To install `rqdatatable` please use `devtools` as follows.

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("WinVector/rqdatatable")
```

