---
title: "Grouped Performance Demonstration"
output: github_document
---

For this example we will look at a per-group maximum calculation over many rows and a few columns. 

First we make the example data.

```{r setup}
set.seed(2020)
nrow <- 1000000
ndcol <- 10

mk_data <- function(nrow, ndcol) {
  d <- data.frame(
    g = sprintf("level_%09g", sample.int(nrow, size = nrow, replace = TRUE)),
    stringsAsFactors = FALSE)
  for(j in seq_len(ndcol)) {
    v <- sprintf("v_%05g", j)
    d[[v]] <- rnorm(nrow)
  }
  return(d)
}

d <- mk_data(nrow, ndcol)
write.csv(d, file = gzfile("d.csv.gz"), quote = FALSE, row.names = FALSE)
vars <- setdiff(colnames(d), 'g')
```

Example processing, `base R`.

```{r baseR}
f_base <- function(d) {
  d_res <- d
  perm <- do.call(order, as.list(d_res[, c('g', vars), drop= FALSE]))
  d_res <- d_res[perm, , drop=FALSE]
  rownames(d_res) <- NULL
  for(v in vars) {
    d_res[[paste0('max_', v)]] = as.numeric(tapply(d_res[[v]], d_res$g, max)[d_res$g])
  }
  d_res
}

res_base <- f_base(d)
```


Example processing, `rqdatatable`.

```{r rqdatatable}
library(rqdatatable)
packageVersion("rquery")
packageVersion("rqdatatable")

ops_rquery <- local_td(d, name = 'd') %.>%
  extend_se(.,
         paste0('max_', vars) %:=% paste0('max(', vars, ')'),
         partitionby = 'g') %.>%
  order_rows(.,
             c('g', vars))

cat(format(ops_rquery))

res_rqdatatable <- d %.>% ops_rquery

stopifnot(isTRUE(all.equal(data.frame(res_base), data.frame(res_rqdatatable))))

knitr::kable(head(res_rqdatatable))

write.csv(res_rqdatatable, file = gzfile("res.csv.gz"), quote = FALSE, row.names = FALSE)
```

Example processing `rquery/db`.

```{r rquerydb}
packageVersion('DBI')
packageVersion('RSQLite')

raw_connection <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
RSQLite::initExtension(raw_connection)
db <- rquery_db_info(
  connection = raw_connection,
  is_dbi = TRUE,
  connection_options = rq_connection_tests(raw_connection))

cat(to_sql(ops_rquery, db))

f_rquery_db <- function(d) {
  rquery::rq_copy_to(db, "d", d, 
                     temporary = TRUE, overwrite = TRUE)
  res <- execute(db, ops_rquery)
  return(res)
}


res_rquery_db <- f_rquery_db(d)

stopifnot(isTRUE(all.equal(data.frame(res_rquery_db), data.frame(res_rqdatatable))))
```



Example processing `data.table`.

```{r data.table}
library(data.table)
packageVersion("data.table")

f_data.table <- function(d) {
  dt <- data.table(d)
  exprs <- paste0('max_', vars, ' = max(', vars, ')')
  stmt <- paste0('dt[, `:=`(', paste(exprs, collapse = ', '), '), by = g]')
  dt <- eval(parse(text=stmt))
  setorderv(dt, c('g', vars))
  return(dt)
}

res_data.table <- f_data.table(d)

stopifnot(isTRUE(all.equal(res_rqdatatable, data.frame(res_data.table))))
```


Example processing, `dplyr`.

```{r dplyr}
library(dplyr)
library(rlang)
packageVersion("dplyr")

exprs <- paste0('max_', vars, ' := max(', vars, ')')
rlang_expr <- eval(parse(text=paste0('exprs(', paste(exprs, collapse = ', '), ')')))
rlang_cols <- syms(c('g', vars))

ops_dplyr <- . %>%
  group_by(g) %>%
  mutate(!!!rlang_expr) %>%
  ungroup() %>%
  arrange(!!!rlang_cols)

res_dplyr <- d %>% ops_dplyr

stopifnot(isTRUE(all.equal(res_rqdatatable, data.frame(res_dplyr))))
```

Example processing, `dbplyr`.

```{r dbplyr}
library(dbplyr)
packageVersion("dplyr")

ops_dbplyr <- tbl(raw_connection, "d") %>%
  group_by(g) %>%
  mutate(!!!rlang_expr) %>%
  ungroup() %>%
  arrange(!!!rlang_cols)

show_query(ops_dbplyr)

f_dbplyr <- function(d) {
  dplyr::copy_to(raw_connection, df=d, name="d", 
                     temporary = TRUE, overwrite = TRUE)
  res <- compute(ops_dbplyr)
  return(res)
}

res_dbplyr <- f_dbplyr(d)

stopifnot(isTRUE(all.equal(data.frame(res_dbplyr), data.frame(res_rqdatatable))))
```


Example processing, `dtplyr`.

```{r dtplyr}
library(dtplyr)
packageVersion("dtplyr")

exprs <- paste0('max_', vars, ' := max(', vars, ')')
rlang_expr <- eval(parse(text=paste0('exprs(', paste(exprs, collapse = ', '), ')')))
rlang_cols <- syms(c('g', vars))
  
ops_dtplyr <- . %>%
  lazy_dt() %>%
  group_by(g) %>%
  mutate(!!!rlang_expr) %>%
  ungroup() %>%
  arrange(!!!rlang_cols) %>%
  as_tibble()

res_dtplyr <- d %>% ops_dtplyr
stopifnot(isTRUE(all.equal(res_rqdatatable, data.frame(res_dtplyr))))
```

```{r microbenchmark}
library(microbenchmark)

microbenchmark(
  base_R = f_base(d),
  data.table = f_data.table(d),
  dplyr = d %>% ops_dplyr,
  dbplyr = f_dbplyr(d),
  dtplyr = d %>% ops_dtplyr,
  rqdatatable = d %.>% ops_rquery,
  rquery_db = f_rquery_db(d),
  times = 5L)
```


Details for a small performance comparison run on 2020-02-26.

Machine was an idle Late 2013 Mac Mini running macOS High Sierra 10.13.6, Processor 2.8 GHz Intel Core i5, Memory 8 GB 1600 MHz DDR3.

```{r rversion}
R.version
```
