---
title: "Speed Up Your R Work"
author: "John Mount"
date: "`r Sys.Date()`"
output: github_document
---

# Introduction

In this note we will show how to speed up work in [`R`](https://www.r-project.org) by partitioning data and process-level parallelization. We will show the technique with three different `R` packages: [`rqdatatable`](https://github.com/WinVector/rqdatatable), [`data.table`](https://CRAN.R-project.org/package=data.table), and [`dplyr`](https://CRAN.R-project.org/package=dplyr).  The methods shown will also work with base-`R` and other packages.

For each of the above packages we speed up work by using [`wrapr::execute_parallel`](https://winvector.github.io/wrapr/reference/execute_parallel.html) which in turn uses [`wrapr::partition_tables`](https://winvector.github.io/wrapr/reference/partition_tables.html) to partition un-related `data.frame` rows and then distributes them to different processors to be executed. [`rqdatatable::ex_data_table_parallel`](https://winvector.github.io/rqdatatable/reference/ex_data_table_parallel.html) conveniently bundles all of these steps together when working with [`rquery`](https://CRAN.R-project.org/package=rquery) pipelines. 

The partitioning is specified by the user preparing a grouping column that tells the system which sets of rows must be kept together in a correct calculation.  We are going to try to demonstrate everything with simple code examples, and minimal discussion.

Keep in mind: unless the pipeline steps have non-trivial cost, the overhead of partitioning and distributing the work may overwhelm any parallel speedup. Also `data.table` itself already seems to exploit some thread-level parallelism (notice user time is greater than elapsed time). That being said, in this note we will demonstrate a synthetic example where computation is expensive due to a blow-up in an intermediate join step.

# Our example

First we set up our execution environment and example (some details: OSX 10.13.4 on a 2.8 GHz Intel Core i5 Mac Mini (Late 2015 model) with 8GB RAM and hybrid disk drive).

```{r ex}
library("rqdatatable")
library("microbenchmark")
library("ggplot2")
library("WVPlots")
suppressPackageStartupMessages(library("dplyr"))

base::date()
R.version.string
parallel::detectCores()
packageVersion("parallel")
packageVersion("rqdatatable")
packageVersion("rquery")
packageVersion("dplyr")

ncore <- parallel::detectCores()
print(ncore)
cl <- parallel::makeCluster(ncore)
print(cl)

set.seed(2362)
mk_example <- function(nkey, nrep, ngroup = 20) {
  keys <- paste0("key_", seq_len(nkey))
  key_group <- sample(as.character(seq_len(ngroup)), 
                      length(keys), replace = TRUE)
  names(key_group) <- keys
  key_table <- data.frame(
    key = rep(keys, nrep),
    stringsAsFactors = FALSE)
  key_table$data <- runif(nrow(key_table))
  instance_table <- data.frame(
    key = rep(keys, nrep),
    stringsAsFactors = FALSE)
  instance_table$id <- seq_len(nrow(instance_table))
  instance_table$info <- runif(nrow(instance_table))
  # groups should be no finer than keys
  key_table$key_group <- key_group[key_table$key]
  instance_table$key_group <- key_group[instance_table$key]
  list(key_table = key_table,
       instance_table = instance_table)
}

dlist <- mk_example(10, 10)
data <- dlist$instance_table
annotation <- dlist$key_table
```

# rquery / rqdatatable

[`rquery`](https://github.com/WinVector/rquery) and [`rqdatatable`](https://github.com/WinVector/rqdatatable) can implement a non-trivial calculation as follows.

```{r rquery1}
# possible data lookup: find rows that
# have lookup data <= info
optree <- local_td(data) %.>%
  natural_join(., 
               local_td(annotation), 
               jointype = "INNER", 
               by = "key") %.>%
  select_rows_nse(., data <= info) %.>%
  pick_top_k(., 
             k = 1,
             partitionby = "id",
             orderby = "data",
             reverse = "data",
             keep_order_column = FALSE) %.>%
  orderby(., "id")
cat(format(optree))

res1 <- ex_data_table(optree)
head(res1)
nrow(res1)
```

Or we could try a theta-join, which reduces production of intermediate rows.

```{r rquery_theta}
# possible data lookup: find rows that
# have lookup data <= info
optree_theta <- local_td(data) %.>%
  theta_join_se(., 
                local_td(annotation), 
                jointype = "INNER", 
                expr = "key == key && info >= data") %.>%
  select_rows_nse(., data <= info) %.>%
  pick_top_k(., 
             k = 1,
             partitionby = "id",
             orderby = "data",
             reverse = "data",
             keep_order_column = FALSE) %.>%
  orderby(., "id")
cat(format(optree_theta))

res_theta <- ex_data_table(optree_theta)
head(res_theta)
nrow(res_theta)
```

And we can execute the operations in parallel.

```{r rqdatatablep1}
parallel::clusterEvalQ(cl, 
                       library("rqdatatable"))
res2 <- ex_data_table_parallel(optree, 
                               "key_group", 
                               cl)
head(res2)
nrow(res2)
```

# data.table

[`data.table`](http://r-datatable.com) can implement the same function.

```{r data_table_f}
library("data.table")
packageVersion("data.table")

# revised function from:
# http://www.win-vector.com/blog/2018/07/speed-up-your-r-work/#comment-66925
data_table_f <- function(data, annotation) {
  #setDT(data, key = c("key","info"))
  #setDT(annotation, key = c("key","data"))
  data <- data.table::as.data.table(data)
  annotation <- data.table::as.data.table(annotation)
  
  joined2 <- data[annotation,
                  on=.(key, info >= data),
                  .(id,
                    key,
                    info = x.info,
                    key_group.x = x.key_group,
                    data = i.data,
                    key_group.y = i.key_group),
                  allow.cartesian=TRUE,
                  nomatch = 0]
  
  setorder(joined2,data)
  joined2[joined2[,.I[.N], keyby = .(id)]$V1]
}
resdt <- data_table_f(data, annotation)
head(resdt)
nrow(resdt)
```

We can also run `data.table` in parallel using [`wrapr::execute_parallel`](https://winvector.github.io/wrapr/reference/execute_parallel.html).

```{r data_table_p}
parallel::clusterEvalQ(cl, library("data.table"))
parallel::clusterExport(cl, "data_table_f")

dt_f <- function(tables_list) {
  data <- tables_list$data
  annotation <- tables_list$annotation
  data_table_f(data, annotation)
}

data_table_parallel_f <- function(data, annotation) {
  respdt <- wrapr::execute_parallel(
    tables = list(data = data, 
                  annotation = annotation),
    f = dt_f,
    partition_column = "key_group",
    cl = cl) %.>%
    data.table::rbindlist(.)
  data.table::setorderv(respdt, cols = "id")
  respdt
}
respdt <- data_table_parallel_f(data, annotation)
head(respdt)
nrow(respdt)
```

# dplyr

[`dplyr`](https://CRAN.R-project.org/package=dplyr) can also implement the example.

```{r dplyr1}
dplyr_pipeline <- function(data, annotation) {
  res <- data %>%
    inner_join(annotation, by = "key") %>%
    filter(data <= info) %>%
    group_by(id) %>%
    arrange(-data) %>%
    mutate(rownum = row_number()) %>%
    ungroup() %>%
    filter(rownum == 1) %>%
    arrange(id)
  res
}

resd <- dplyr_pipeline(data, annotation)
head(resd)
nrow(resd)
```

And we can use [`wrapr::execute_parallel`](https://winvector.github.io/wrapr/reference/execute_parallel.html) to parallelize the `dplyr` solution.

```{r dplyr_wp}
parallel::clusterEvalQ(cl, library("dplyr"))
parallel::clusterExport(cl, "dplyr_pipeline")

dplyr_f <- function(tables_list) {
  data <- tables_list$data
  annotation <- tables_list$annotation
  dplyr_pipeline(data, annotation)
}

dplyr_parallel_f <- function(data, annotation) {
  respdt <- wrapr::execute_parallel(
    tables = list(data = data, 
                  annotation = annotation),
    f = dplyr_f,
    partition_column = "key_group",
    cl = cl) %>%
    dplyr::bind_rows() %>%
    arrange(id)
}
respdplyr <- dplyr_parallel_f(data, annotation)
head(respdplyr)
nrow(respdplyr)
```

# Benchmark

We can benchmark the various realizations.

```{r, timings}
dlist <- mk_example(300, 300)
data <- dlist$instance_table
annotation <- dlist$key_table

timings <- microbenchmark(
  data_table_parallel = 
    nrow(data_table_parallel_f(data, annotation)),
  data_table = nrow(data_table_f(data, annotation)),
  
  rqdatatable_parallel = 
    nrow(ex_data_table_parallel(optree, "key_group", cl)),
  rqdatatable = nrow(ex_data_table(optree)),

  rqdatatable_theta_parallel = 
    nrow(ex_data_table_parallel(optree_theta, "key_group", cl)),
  rqdatatable_theta = nrow(ex_data_table(optree_theta)),
  
  dplyr_parallel = 
    nrow(dplyr_parallel_f(data, annotation)),
  dplyr = nrow(dplyr_pipeline(data, annotation)),
  
  times = 10L)

saveRDS(timings, "Parallel_rqdatatable_timings.RDS")
```

# Conclusion


```{r present}
print(timings)

# autoplot(timings)

timings <- as.data.frame(timings)
timings$seconds <- timings$time/1e+9

ScatterBoxPlotH(timings, 
                xvar = "seconds", yvar = "expr", 
                title="task duration by method")
```

In these timings `data.table` is by far the fastest.  Part of it is the faster nature of `data.table`, and another contribution is `data.table`'s non-equi join avoids a lot of expense (which is why theta-style joins are in fact interesting).

A reason `dplyr` sees greater speedup relative to its own non-parallel implementation (yet does not beat `data.table`) is that `data.table` starts already multi-threaded, so `data.table` is exploiting some parallelism even before we added the process level parallelism (and hence sees less of a speed up, though it is fastest).

`rquery` pipelines [exhibit superior performance on big data systems](https://github.com/WinVector/rquery/blob/master/extras/PerfTest.md) (Spark, PostgreSQL, Amazon Redshift, and hopefully soon Google bigquery), and `rqdatatable` supplies [a very good in-memory implementation of the `rquery` system](http://www.win-vector.com/blog/2018/06/rqdatatable-rquery-powered-by-data-table/) based on `data.table`. `rquery` also speeds up solution development by supplying higher order operators and early debugging features.

In this note we have demonstrated simple procedures to reliably parallelize any of `rqdatatable`, `data.table`, or `dplyr`.

Note: we did not include alternatives such as `multidplyr` or `dtplyr` in the timings, as they did not appear to work on this example.


###################

# Materials

The original rendering of this article can be found [here](https://github.com/WinVector/rqdatatable/blob/master/extras/Parallel_rqdatatable.md), source code [here](https://github.com/WinVector/rqdatatable/blob/master/extras/Parallel_rqdatatable.Rmd), and raw timings [here](https://github.com/WinVector/rqdatatable/blob/master/extras/Parallel_rqdatatable_timings.RDS).


# multidplyr

[`multidplyr`](https://github.com/hadley/multidplyr) does not appear to work on this example,
so we could not include it in the timings.

```{r multidplyr, error=TRUE}
# devtools::install_github("hadley/multidplyr")
library("multidplyr") # https://github.com/hadley/multidplyr
packageVersion("multidplyr")
multidplyr::set_default_cluster(cl)

# example similar to https://github.com/hadley/multidplyr/blob/master/vignettes/multidplyr.Rmd
class(data)
datap <- multidplyr::partition(data, key_group)
head(datap)
class(datap)

class(annotation)
annotationp <- multidplyr::partition(annotation, key_group)
head(annotationp)
class(annotationp)

dplyr_pipeline(datap, annotationp) %>%
  collect()
```

# dtplyr

[`dtplyr`](https://CRAN.R-project.org/package=dtplyr) does not appear to work on this example, so we could not include it in the timings.

```{r dtplyr, error=TRUE}
library("data.table")
library("dtplyr") #  https://CRAN.R-project.org/package=dtplyr
packageVersion("dtplyr")

class(data)
datadt <- data.table::as.data.table(data)
head(datadt)
class(datadt)

class(annotation)
annotationdt <- data.table::as.data.table(annotation)
head(annotationdt)
class(annotationdt)

dplyr_pipeline(datadt, annotationdt)
```

# clean up

```{r cleanup}
parallel::stopCluster(cl)
rm(list = "cl")
```
