---
title: "Speeding Up R Calculations with Parallelization"
author: "John Mount"
date: "`r Sys.Date()`"
output: github_document
---

# Introduction

In this note we will demonstrate speeding up calculations by partitioning data and process-level parallelization. 

We will show how to speed up calculations with parallelization using [`rqdatatable`](https://github.com/WinVector/rqdatatable), [`data.table`](https://CRAN.R-project.org/package=data.table), or [`dplyr`](https://CRAN.R-project.org/package=dplyr).  For each of these packages the parallelization becomes possible when we use [`wrapr::execute_parallel`](https://winvector.github.io/wrapr/reference/execute_parallel.html) to partition un-related `data.frame` rows to be distributed to different processors.

However, unless the pipeline steps have non-trivial cost,
the overhead of partitioning and distributing the work may overwhelm any parallel speedup.
Also `data.table` itself already seems to exploit some thread-level parallelism (notice user time is greater than elapsed time). 
That being said, we can test an synthetic example where computation is expensive due to a blow-up in an intermediate join step.

# Our example

First we set up our execution environment and example (some details: OSX 10.13.4 on a 2.8 GHz Intel Core i5 Mac Mini (Late 2015 model) with 8GB ram and hybrid disk drive).

```{r ex}
library("rqdatatable")
library("microbenchmark")
library("ggplot2")
library("WVPlots")
suppressPackageStartupMessages(library("dplyr"))

base::date()
R.version.string
parallel::detectCores()
packageVersion("parallel")
packageVersion("rqdatatable")
packageVersion("rquery")
packageVersion("dplyr")

ncore <- parallel::detectCores()
print(ncore)
cl <- parallel::makeCluster(ncore)
print(cl)

set.seed(2362)
mk_example <- function(nkey, nrep, ngroup = 20) {
  keys <- paste0("key_", seq_len(nkey))
  key_group <- sample(as.character(seq_len(ngroup)), 
                      length(keys), replace = TRUE)
  names(key_group) <- keys
  key_table <- data.frame(
    key = rep(keys, nrep),
    stringsAsFactors = FALSE)
  key_table$data <- runif(nrow(key_table))
  instance_table <- data.frame(
    key = rep(keys, nrep),
    stringsAsFactors = FALSE)
  instance_table$id <- seq_len(nrow(instance_table))
  instance_table$info <- runif(nrow(instance_table))
  # groups should be no finer than keys
  key_table$key_group <- key_group[key_table$key]
  instance_table$key_group <- key_group[instance_table$key]
  list(key_table = key_table,
       instance_table = instance_table)
}

dlist <- mk_example(10, 10)
data <- dlist$instance_table
annotation <- dlist$key_table
```

# rquery / rqdatatable

[`rquery`](https://github.com/WinVector/rquery) and [`rqdatatable`](https://github.com/WinVector/rqdatatable) can implement a non-trivial calculation as follows.

```{r rquery1}
# possible data lookup: find rows that
# have lookup data <= info
optree <- local_td(data) %.>%
  natural_join(., 
               local_td(annotation), 
               jointype = "INNER", 
               by = "key") %.>%
  select_rows_nse(., data <= info) %.>%
  pick_top_k(., 
             k = 1,
             partitionby = "id",
             orderby = "data",
             reverse = "data",
             keep_order_column = FALSE) %.>%
  orderby(., "id")
cat(format(optree))

res1 <- ex_data_table(optree)
head(res1)
nrow(res1)
```

And we can execute the operations in parallel.

```{r rqdatatablep1}
parallel::clusterEvalQ(cl, 
                       library("rqdatatable"))
res2 <- ex_data_table_parallel(optree, 
                               "key_group", 
                               cl)
head(res2)
nrow(res2)
```

# data.table

[`data.table`](http://r-datatable.com) can implement the same function.

```{r data_table_f}
library("data.table")
packageVersion("data.table")

data_table_f <- function(data, annotation) {
  data <- data.table::as.data.table(data)
  annotation <- data.table::as.data.table(annotation)
  joined <- merge(data, annotation, 
                  by = "key", 
                  all=FALSE, 
                  allow.cartesian=TRUE)
  joined <- joined[joined$data <= joined$info, ]
  data.table::setorderv(joined, cols = "data")
  joined <- joined[, .SD[.N], id]
  data.table::setorderv(joined, cols = "id")
}
resdt <- data_table_f(data, annotation)
head(resdt)
nrow(resdt)
```

We can also run `data.table` in parallel using [`wrapr::execute_parallel`](https://winvector.github.io/wrapr/reference/execute_parallel.html).

```{r data_table_p}
parallel::clusterEvalQ(cl, library("data.table"))
parallel::clusterExport(cl, "data_table_f")

dt_f <- function(tables_list) {
  data <- tables_list$data
  annotation <- tables_list$annotation
  data_table_f(data, annotation)
}

data_table_parallel_f <- function(data, annotation) {
  respdt <- wrapr::execute_parallel(
    tables = list(data = data, 
                  annotation = annotation),
    f = dt_f,
    partition_column = "key_group",
    cl = cl) %.>%
    data.table::rbindlist(.)
  data.table::setorderv(respdt, cols = "id")
  respdt
}
respdt <- data_table_parallel_f(data, annotation)
head(respdt)
nrow(respdt)
```

# dplyr

[`dplyr`](https://CRAN.R-project.org/package=dplyr) can also implement the example.

```{r dplyr1}
dplyr_pipeline <- function(data, annotation) {
  res <- data %>%
    inner_join(annotation, by = "key") %>%
    filter(data <= info) %>%
    group_by(id) %>%
    arrange(-data) %>%
    mutate(rownum = row_number()) %>%
    ungroup() %>%
    filter(rownum == 1) %>%
    arrange(id)
  res
}

resd <- dplyr_pipeline(data, annotation)
head(resd)
nrow(resd)
```

And we can use [`wrapr::execute_parallel`](https://winvector.github.io/wrapr/reference/execute_parallel.html) to
also parallelize the `dplyr` solution.

```{r dplyr_wp}
parallel::clusterEvalQ(cl, library("dplyr"))
parallel::clusterExport(cl, "dplyr_pipeline")

dplyr_f <- function(tables_list) {
  data <- tables_list$data
  annotation <- tables_list$annotation
  dplyr_pipeline(data, annotation)
}

dplyr_parallel_f <- function(data, annotation) {
  respdt <- wrapr::execute_parallel(
    tables = list(data = data, 
                  annotation = annotation),
    f = dplyr_f,
    partition_column = "key_group",
    cl = cl) %>%
    dplyr::bind_rows() %>%
    arrange(id)
}
respdplyr <- dplyr_parallel_f(data, annotation)
head(respdplyr)
nrow(respdplyr)
```

# Benchmark

We can benchmark the various realizations.

```{r, timings}
dlist <- mk_example(300, 300)
data <- dlist$instance_table
annotation <- dlist$key_table

timings <- microbenchmark(
  data_table_parallel = 
    nrow(data_table_parallel_f(data, annotation)),
  data_table = nrow(data_table_f(data, annotation)),
  rqdatatable_parallel = 
    nrow(ex_data_table_parallel(optree, "key_group", cl)),
  rqdatatable = nrow(ex_data_table(optree)),
  dplyr_parallel = 
    nrow(dplyr_parallel_f(data, annotation)),
  dplyr = nrow(dplyr_pipeline(data, annotation)),
  times = 10L)

saveRDS(timings, "Parallel_rqdatatable_timings.RDS")
```

# Conclusion


```{r present}
print(timings)

# autoplot(timings)

timings <- as.data.frame(timings)
timings$seconds <- timings$time/1e+9

ScatterBoxPlotH(timings, 
                xvar = "seconds", yvar = "expr", 
                title="task duration by method")
```

Parallelized `data.table` is the fastest, followed by parallelized `dplyr` and parallelized `rqdatatable`.  
The non-paraellized run times are in a similar order.  A reason `dplyr` sees greater speedup relative
to its own non-parallel implementation is that `data.table` starts already multi-threaded, so it is exploring
some parallelism even before we added the fork-style parallelism.  We did not include variations such as `multidplyr` or `dtplyr` 
in the timings, as they did not appear to work.


###################

# Materials

The original rendering of this article can be found [here](https://github.com/WinVector/rqdatatable/blob/master/extras/Parallel_rqdatatable.md), source code [here](https://github.com/WinVector/rqdatatable/blob/master/extras/Parallel_rqdatatable.Rmd), and raw timings [here](https://github.com/WinVector/rqdatatable/blob/master/extras/Parallel_rqdatatable_timings.RDS).

# Speculation

`rqdatatable`'s minor performance regression relative to `datatable` I believe is from `rqdatatable`'s ranking strategy (something we will likely tune later, already [usually `rqdatatable` is competitive with `data.table` and actually quite fast](https://github.com/WinVector/rquery/blob/master/extras/data_table_replot.md)).

# multidplyr

[`multidplyr`](https://github.com/hadley/multidplyr) does not appear to work on this example,
so we could not include it in the timings.

```{r multidplyr, error=TRUE}
library("multidplyr") # https://github.com/hadley/multidplyr
packageVersion("multidplyr")
multidplyr::set_default_cluster(cl)

# example similar to https://github.com/hadley/multidplyr/blob/master/vignettes/multidplyr.Rmd
class(data)
datap <- multidplyr::partition(data, key_group)
head(datap)
class(datap)

class(annotation)
annotationp <- multidplyr::partition(annotation, key_group)
head(annotationp)
class(annotationp)

dplyr_pipeline(datap, annotationp) %>%
  collect()
```

# dtplyr

[`dtplyr`](https://CRAN.R-project.org/package=dtplyr) does not appear to work on this example,
so we could not include it in the timings.

```{r dtplyr, error=TRUE}
library("data.table")
library("dtplyr") #  https://CRAN.R-project.org/package=dtplyr
packageVersion("dtplyr")

class(data)
datadt <- data.table::as.data.table(data)
head(datadt)
class(datadt)

class(annotation)
annotationdt <- data.table::as.data.table(annotation)
head(annotationdt)
class(annotationdt)

dplyr_pipeline(datadt, annotationdt)
```

# clean up

```{r cleanup}
parallel::stopCluster(cl)
rm(list = "cl")
```
